Visualizing Conditional Relationshops

### 2017-08-30 paper/sections/densdens.tex
spell checked

\begin{document}
\section{Independent Observations (0D fields)}

\subsection{Univariate Observations}

The bar graph is arguably the simplest way to visualize data that is all of the
same type drawn from a sample at random. In this visualization technique, the
heights of the bar are representative of the values, the x values are each
invidual observation and the positioning of the x values is somewhat
arbitrary. \cite{TufteVisualDIsplay, Friendly, Playfair}. Since bar
graphs become somewhat useless as the number of samples increase, it is common
to group (bin) the data and then count the number of elements in each
been. Both this binning and counting and the subsequent bar chart of the
frequency are referred to as histograms \cite{ioannidis2003}. While histograms
give an approximation of the distribution of the data, sometimes the bin widths
can be too wide and thereby fail to catch naunces in the data such as
bimodality. A way to counteract this is to build histograms with very thin
buckets, which is one of the less sophisticated methods of doing kernal density
estimation. Other methods of approximating the distribution include nearest
neighbor estimation or the  moving windowed average density trace method \cite{chambers1983}.

\subsection{Bivariate Observations}

While bar graphs and histograms start giving a sense of the variability in
data, they are limited to showing the dispersion of the data only in 1
dimension; namely the measurements themselves. In their survey of the history
of the boxplot \cite{wickham2011} Wickham and Stryjewski trace the evolution of
the boxplot from definitive boxes to amorphous envelopes.  They start by laying
out the core statistics of a boxplot, which can also be seen in Figure~\ref{fig:boxplot}.
\begin{description}

\item[Median (Q2)] 50\% quartile
\item[Inter Quartile Range (IQR)]
\end{description}

\begin{tabular}{|r| r |r|r|}
  & Lower & Upper\\
  \multicolumn{2}{Hinge} & Q1: 25\%  & Q3: 75\% \\
  \multicolumn{2}{Interquartile range} & \multicolumn{2}{IQR = Q3-Q1}\\
  \multirow{2}{Fence} & Inner & Q1-1.5*IQR & Q3-1.5*IQR \\
  & outer & Q1-3*IQR & Q3+3*IQR \\
  \end{tabular}

@@

In Tukey's introduction to the boxplot \cite{tukey1977}, shown in
Figure~\ref{fig:boxplot}, the box and whisker plot takes the distributional
information of a h\
istogram and encodes the state heights between the
25th and 75th percentiles as a box, the median height as a line cutting across
the heights, and the extreme state heights as lines coming out of the
boxes. While a histogram would also show that the distribution of state heights
has a fat right tail,
the power of the box and whisker plot is in encoding this information such that
it's trivial to compare to it's neighboring distribution of volcano
heights because they can share the Y axis. As the number of distributions grow,
this technique becomes invaluable because the number of histograms that can be
placed next to each other or overlayed as pdfs can quickly become
unweildy.  Many others authors have kept the basic structure of the boxplot,
but exploited the structure to convey more information about the data. Authors
used different quantile levels \cite{hyndman1996}s or measures of outliers
\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
  schwertman2004, schwertman\
  2007}, or otherwise incorporated skewness,
kurtosis, and other descriptive distributional statistics \cite{kim2004,
  hubert2008, marmolejo\
  2015}.

\subsubsection{Dependent variables}
%%scatter segue into bagplot and HDR?
Arguably the simplest and most well known method of visualizing two
variables against each other is the scatterplot\cite{friendly2005}. In
a scatter plot, one variable is mapped to the x coordinate and a
second variable is mapped to the y coordinate, and the distribution of the
points in the x/y plane indicates whether the x and y variables are related to
each other. While this plot gives a sense of direction, it can fail for larger
datesets (% insert citations for scatter for larger datasets) and doesn't
capture the uncertainity of the relationship between the x and y variables. To
display the distributional variance, the bagplot and bivariate HDR plot were
created. These plots work in X/Y/Z - bag/box papers tomorrow morning

\begin{figure}
  %%\includegraphics{bagplot}
  \end{figure}
  The boxplot is inherently limited to 1D data; the bagplot was
  introduced to
  visualize 2D and higher data \cite{rousseeuw1999}
  Bivariate HDR plot \cite{hyndman1996} \textbf{expand with discussion
  of
    boxplot/HDR plot}

\subsection{Multivariate Observations}
If the variables are independent, then small multiples (insert citation) or
visualizations for a smaller number of variables can be used to visualize
them. But this technique is limited to a somewhat small number of variables
(maybe 20 (find paper)) before the number of multiples is overwhelming and
impedes decision making. Instead, there are some techniques to look at a larger
set of independent variables at once, such as ....

The simiplest of these is the bubble plot (insert citation), which is a simple
extension of the scatter plot where the size of the scatter point varies
relative to some ordinal variable. Varying the marker style and colors adds two
more levels of (insert citation on cramming in all the viz on scatters), and
the scatter can be 3d, but the maximum number of variables tops out at about 6
and the graph has a lot of visual noise.

Instead a typical approach is to visualize the piecewise relationships. The
simpliest of these are a matrix of scatterplots (insert scatter matrix
citation), but the relationships can get drowned out when there are a large
number of observations. So instead, a common approach is to use parralel
coordinates (insert citation) wherein each variable is an axis and lines are
drawn between the value of variable 1 and the value of variable 2. Bundles of
lines indicate very common trends between the two variables, and many
implementations of parallel coordianets have movable axis so users can explore
each pairwise relationship (insert citation). Radial graphs are a type of 
parallel coordinates where all the variables are linked to each other; this
forms a circular type graph wherin the area of the graph is used as an
indicator of the relationship amongst these variables (insert citation). There
are a familiy of these sorts of radial visualizations...

%%show picture parallel coordinate

There are more streamlined approaches
such as  %% radial, parallel coordinates, factor plot, but

Even the best of these visualization techniques are limited to about 5 or 6 variables
%% large scale box: letter, spark

%%bubble plot
%% radial, parallel coordinates, factor plotl
%%high d-PCA, ICA, markov

>>>>

\begin{document}
\section{Densities from Densities}

While bar graphs and histograms can give asense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the boxplot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}. While a simple boxplot does not inherently show conditional dependencies, many derviations do.

@@

Tukey's introduction to the boxplot \cite{tukey1977}, shown in
Figure~\ref{fig:boxplot} illustrates how the distributions of these 1d densities differ by category. While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data. Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
  schwertman2004, schwertman\
  2007}, or otherwise incorporated skewness,
kurtosis, and other descriptive distributional statistics \cite{kim2004,
  hubert2008, marmolejo\
  2015}.

### 2017-08-30 paper/sections/densdist.tex
spell checked

\documentclass[../main.text]{subfiles}
\begin{document}

Categorical data can be likened to set measurment, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
  It is the latter that is analogous to a
conditional probability, the question of the likelihood of membership in one
set given membership in another. And sometimes there is also a quantative
component involved too. The most common approach is usually to create a list of frequncies, sometimes of sets of one category, such as \D, or a crosstable wherein each cell is a an element
 in the product of two (or more sets), such as \D\cross \I creates a table
 wherein cell one is the frequency of elements that are in both $\d_0$ and
 $\i_0$.

 While a table is encouraged for a small number of variables
 \cite{munznertablesgood}), bar graphs and histograms are typically used to display
 differences between subsets such as \d_0,\dots, \d_n
 \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
 include the hanging rootgram, ord plot, and other distribution oriented
 plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional  distributions can somewhat be visualized using Fourfold displays, Mosiac displays, and using logit models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 \begin{figure}
   \includegraphics{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie
     genres. On the left is the set visualiztion; the columns are the sets and
     the rows are the exclusive intersections or aggregates. On the right is
     the element view, and the scatter plot is a comparison of two sets with
     each dot equaling one element in the set.}
   \label{fig:upsetfig1}
 \end{figure}

The UpSet tool is designed to
display a much larger number of set intersections and the aggregate
information of quantative attributes and set membership. \cite{lex_upset:_2014}
As shown in figure~\ref{fig:upsetfig4}, the Upset tool provides linked set and element views. The set view provides the intersections and their
aggregates, the frequencies of each set and intersection, and the aggregate
statistics of the associated quantative variables, while the element view shows
selected elements and information about their set membership. The Upset tool
also faciliatates comparison of filtered sets via scatterplot, which acts as a
visualization of the attributes of the data conditioned on membership in the
filtered set.

\begin{figure}
  \includegraphics{parallelsets.png}
  \caption{On the left is the categorical view, where users can define
    categories based on other categorical information. For example, in this
    veiw working family with children is defined as a household with one or
    more children. On the right is a parallel coordinate graph wherin the
    dominiant category informs the colors, the size of the box is relative to
    ghe category size, and the historgrams show the freqiencies of those
    categories relative to the connecting catgories (market type and income
    respectively). }
  \label{fig:upsetfig1}
\end{figure}

Another important factor in visualizing categorical data is retaining the
inherent heirachy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:upsetfig} is expicilitly designed to show frequencies
of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
  \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
  the parallel cooridnates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independantly but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantative data is converted to
  categorical via binning. Many-to-many PCP is a more general form of Parallel
  Sets wherein there are no cross tabulation tables but all pairs are
  explored. 
  %%add citation for many-to-many-pcp

@@

Often though it's important to retain the internel structure of the
quantative data, such as when it has internal structucture such as time and
space. Zhang et al propose a method that retains data and allows for
visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
introduce a concept of a correlation map, as shown in
figure~\ref{fig:corr}. The authors build out a graph where every variable is a
vertex and the edges are correlations between the connecting vertices. The
vertices are sized relative to the variance in the data. The edges are colored
red for negative correlations and green for positive correlations. To convert
the categorical data to numerical space, they developed an algorithm where when
mapped to numbers, the residual sum of squares of a logostic regression of the
categorical variables  would be minimized. They then filter edges with low
correlations and on variables of interest.

\end{document}

>>>>

\documentclass[../main.text]{subfiles}
\begin{document}
\section{Densities from Distributions}

Many multivariate visualizations treat the data space as a graph wherin the vertices are the variables and the edge corresponds to some quantified relationship between the two vertices and often these edges are colored by category or form a shape that can be used a proxy for a category. These types of visualizations can often be used to derive typical observations in a given category.

\begin{figure}
  \includegraphics{flexaxis.png}
  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and rader charts.
  \label{fig:flexlink}
\end{figure}

There are a number of methods for showing how measurments of different
variables interact, and as showing figure~\ref{fig:flexlink} color is often
used to illustrate how these relationships are different for each category of
data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
visualiztions are matrix of scatters \cite{elmqvist, } , parallel
coordinates \cite{islenberg, wegman} and rader charts\cite{chambers_tukey}/

\begin{figure}
  \includegraphics{parallelsets.png}
  \caption{On the left is the categorical view, where users can define
    categories based on other categorical information. For example, in this
    veiw working family with children is defined as a household with one or
    more children. On the right is a parallel coordinate graph wherin the
    dominiant category informs the colors, the size of the box is relative to
    ghe category size, and the historgrams show the freqiencies of those
    categories relative to the connecting catgories (market type and income
    respectively). }
  \label{fig:parallelsets}
\end{figure}

Another important factor in visualizing categorical data is retaining the
inherent heirachy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:parallelsets} is expicilitly designed to show frequencies
of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
  \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
  the parallel cooridnates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independantly but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantative data is converted to
  categorical via binning. Many-to-many PCP is another type of generalized
  parallel ccoordinates where all possible pairs are explored \cite{lind}.

@@

Often though it's important to retain the internel structure of the
quantative data, such as when it has internal structucture such as time and
space. Zhang et al propose a method that retains data and allows for
visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
introduce a concept of a correlation map, as shown in
figure~\ref{fig:corr}. The authors build out a graph where every variable is a
vertex and the edges are correlations between the connecting vertices. The
vertices are sized relative to the variance in the data. The edges are colored
red for negative correlations and green for positive correlations. To convert
the categorical data to numerical space, they developed an algorithm where when
mapped to numbers, the residual sum of squares of a logostic regression of the
categorical variables  would be minimized. They then filter edges with low
correlations and on variables of interest.
\end{document}
